---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r}
library(dplyr)
library(ggplot2)
```


```{r}
Sys.setlocale("LC_TIME", "en_US")
stepdata=read.csv("./activity.csv")
stepdata$date=as.Date(stepdata$date)
stepdata$weekday=weekdays(stepdata$date)
head(stepdata)
```

## Histogram of the total number of steps taken each day

```{r}
steps_per_day = stepdata %>% group_by(date) %>% summarise(step.by.date=sum(steps,na.rm=TRUE))
head(steps_per_day )
```

```{r}
hist(steps_per_day$step.by.date, main = "Total number of steps taken per day", xlab = "Total steps taken per day", col = "darkblue", ylim = c(0,20), breaks = seq(0,25000, by=2500))

```

## What is mean total number of steps taken per day?
```{r}
mean_steps=mean(steps_per_day$step.by.date)
median_steps=median(steps_per_day$step.by.date)
cat("Mean =",mean_steps, "| Median =", median_steps)
```

## What is the average daily activity pattern?
Time series plot for daily step taken
```{r}
mean_steps_by_date=stepdata %>% group_by(date) %>% summarise(mean.steps=mean(steps,na.rm=TRUE))
with(mean_steps_by_date,plot(date,mean.steps,type="l",lwd=2))
```


Time series plot for average daily pattern by intervals
```{r}
mean_steps_by_interval=stepdata %>% group_by(interval) %>% summarise(mean.steps=mean(steps,na.rm=TRUE))
head(mean_steps_by_interval)
```

```{r}
with(mean_steps_by_interval,plot(interval,mean.steps,type="l",xlab = "Interval", ylab = "Average number of steps", main = "Average number of steps per interval"))
```

5-minute interval, on average across all the days in the dataset, contains the maximum number of steps
```{r}
mean_steps_by_interval[which.max(mean_steps_by_interval$mean.steps),]$interval
```


## Imputing missing values
```{r}
na.values=sum(is.na(stepdata$steps))
cat("Total numbers of NAs is ",na.values,", this acounts for ",na.values/nrow(stepdata
                                                                              )*100, "% of the dataset")
```
The strategy is to fill all of the NAs by the average steps of the correspondent intervals


```{r}
filled_NA=left_join(stepdata,mean_steps_by_interval,by="interval")
filled_NA$steps=ifelse(is.na(filled_NA$steps), yes=filled_NA$mean.steps,no=filled_NA$steps)
head(filled_NA)
fill_NA_sum= filled_NA %>% group_by(date) %>% summarise(total.steps=sum(steps))
```

Histogram of the total number of steps taken each day after missing values are imputed
```{r}
par(mfrow = c(1, 2))
hist(steps_per_day$step.by.date,main="total steps each day in the \noriginal data",breaks = seq(0,25000, by=2500),xlab = "Total steps by date")
hist(fill_NA_sum$total.steps,main="total steps each day in the \nclean data",breaks = seq(0,25000, by=2500),xlab = "Total steps by date")
```


## Are there differences in activity patterns between weekdays and weekends?
```{r}
filled_NA$datetype=ifelse((filled_NA$weekday %in% c("Saturday","Sunday")),"weekend","weekday")
pattern=filled_NA %>% group_by(interval, datetype) %>% summarise(average.steps=mean(steps))
```

Compare the average steps for weekday and weekend

```{r  echo = TRUE }
g=ggplot(pattern,aes(interval,average.steps,col=datetype))
g+geom_line()+facet_wrap(.~datetype,ncol=1, nrow=2) + labs(title = "Average daily steps by date type", x = "Interval", y = "Average number of steps")
```

